---
title: "Hierarchical kNN Classification"
author: "cytoGateR authors"
vignette: >
  %\VignetteIndexEntry{4. Hierarchical kNN Classification}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Overview

Flat classifiers predict among all cell types in a single step. Hierarchical
k-nearest-neighbour (kNN) classification instead organizes reference cell types
into a tree and predicts one branch at a time. Each internal node separates the
cell types on its left branch from those on its right branch. A cell moves
recursively through the hierarchy until it reaches a terminal cell type or
fails the confidence and agreement requirements.

This vignette demonstrates the complete hierarchical kNN workflow on one
complete image from `cytoGateR_example`.

```{r}
library(cytoGateR)
data("cytoGateR_example", package = "cytoGateR")

image_ids <- SummarizedExperiment::colData(cytoGateR_example)$image_name
spe <- cytoGateR_example[, image_ids == image_ids[1L]]
spe
```

# Prepare reference labels

The packaged `core_group` column contains high-confidence labels produced by
tree gating. Cells outside the high-confidence reference are marked
`"Unassigned"`. The hierarchy builders use `"Unknown"` for cells that should
not contribute to the reference, so the labels are harmonized before fitting.

```{r}
cd <- SummarizedExperiment::colData(spe)
cd$hier_reference_label <- as.character(cd$core_group)
cd$hier_reference_label[cd$hier_reference_label == "Unassigned"] <- "Unknown"
SummarizedExperiment::colData(spe) <- cd

table(SummarizedExperiment::colData(spe)$hier_reference_label)
```

# Build the lineage hierarchy

`build_lineage_hierarchy()` calculates the average marker-expression profile
for each reference cell type and hierarchically clusters those profiles.

```{r}
hc_tree <- build_lineage_hierarchy(
  spe,
  label_col = "hier_reference_label",
  assay_name = "exprs"
)

hc_tree$labels
plot(hc_tree, main = "Hierarchy of reference cell types")
```

# Build node-specific references

Each internal node receives its own binary training reference. Labels are
converted to `"Left"` and `"Right"` according to the branches of `hc_tree`.

Setting `top_n = NULL` uses all available markers at each node. For larger
panels, `marker_stats` and `top_n` can be supplied to retain only the most
informative markers for each split.

```{r}
hier_reference <- build_hierarchical_reference(
  spe,
  hc_tree,
  label_col = "hier_reference_label",
  top_n = NULL,
  assay_name = "exprs",
  unknown_label = "Unknown"
)

names(hier_reference)
```

Inspect the root-node reference:

```{r}
root_name <- paste0("Node_", nrow(hc_tree$merge))
root_reference <- hier_reference[[root_name]]

dim(root_reference$train_data)
table(root_reference$train_labels)
root_reference$left_members
root_reference$right_members
```

# Run recursive hierarchical kNN

At each node, `predict_hierarchical_knn_recursive()` creates an ensemble of
weighted kNN predictions. A prediction is accepted only when both its average
probability and ensemble agreement pass their thresholds.

The vignette uses one distance method and one repeat for speed. Production
analyses can use multiple distance methods and additional repeats for a more
stable ensemble.

```{r}
set.seed(1)
spe <- predict_hierarchical_knn_recursive(
  spe,
  hier_reference,
  hc_tree,
  assay_name = "exprs",
  threshold = 0.5,
  agreement_threshold = 0.5,
  k = 3,
  repeats = 1,
  dist_methods = "pearson",
  BPPARAM = BiocParallel::SerialParam(),
  out_col = "hier_label",
  chunk_size = 1000,
  unassigned_label = "Unassigned"
)
```

# Inspect hierarchical predictions

```{r}
table(SummarizedExperiment::colData(spe)$hier_label)
```

The predictions can be compared with labels already stored in the demo object:

```{r}
label_confusion_matrix(
  spe,
  label_col1 = "knn_label",
  label_col2 = "hier_label"
)
```

```{r}
label_agreement_rates(
  spe,
  label_col1 = "knn_label",
  label_col2 = "hier_label"
)
```

Finally, display the hierarchical predictions in spatial coordinates:

```{r}
plot_labelled_cells(
  spe,
  col_label = "hier_label",
  image_col = "image_name"
)
```

# Production settings

A larger ensemble can combine correlation- and angle-based similarity measures:

```{r}
#| eval: false
spe <- predict_hierarchical_knn_recursive(
  spe,
  hier_reference,
  hc_tree,
  threshold = 0.7,
  agreement_threshold = 0.8,
  k = 5,
  repeats = 5,
  dist_methods = c("pearson", "cosine"),
  BPPARAM = BiocParallel::SerialParam()
)
```

# Session information

```{r}
sessionInfo()
```
