---
title: "Introduction to cytoGateR"
author: "cytoGateR authors"
vignette: >
  %\VignetteIndexEntry{1. Introduction to cytoGateR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Introduction

`cytoGateR` provides marker-aware cell-type annotation for spatial protein
imaging data. It combines interpretable marker gating with supervised
prediction of cells that cannot be confidently gated. The package operates on
`SpatialExperiment` and `SummarizedExperiment` objects and stores labels,
probabilities, and uncertainty measurements in standard Bioconductor
containers.

The package is designed for imaging assays with a relatively small protein
panel, where marker knowledge remains important and annotation decisions should
be inspectable. The main workflow consists of:

1. defining positive and negative markers for candidate cell types;
2. estimating cell-type probabilities with tree or soft gating;
3. selecting high-confidence reference cells;
4. predicting the remaining cells with a supervised classifier; and
5. visualising labels, probabilities, and uncertainty.

# Installation

Install the package from Bioconductor:

```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("cytoGateR")
```

Load the package:

```{r}
library(cytoGateR)
```

# Example data

The package includes a compact `SpatialExperiment` derived from the IMMUcan
2022 Cancer Example dataset. It contains normalized marker expression, spatial
coordinates, tree-gating results, high-confidence reference labels, and
weighted k-nearest-neighbour (kNN) predictions.

```{r}
data("cytoGateR_example", package = "cytoGateR")
cytoGateR_example
```

The object contains 40 protein markers:

```{r}
head(rownames(cytoGateR_example))
SummarizedExperiment::assayNames(cytoGateR_example)
```

Sample and image identifiers are stored in `colData()`:

```{r}
cd <- SummarizedExperiment::colData(cytoGateR_example)
table(cd$sample_id)
table(cd$image_name)
```

Spatial coordinates are available through the standard
`SpatialExperiment::spatialCoords()` accessor:

```{r}
head(SpatialExperiment::spatialCoords(cytoGateR_example))
```

# Stored annotation outputs

Tree-gating probabilities use the prefix `P_`, while the complete kNN
probability matrix uses the prefix `KNN_P_`.

```{r}
tree_probability_columns <- grep("^P_", names(cd), value = TRUE)
knn_probability_columns <- grep("^KNN_P_", names(cd), value = TRUE)

tree_probability_columns
knn_probability_columns
```

The stored kNN probabilities form a complete probability distribution for each
cell:

```{r}
knn_probabilities <- as.matrix(cd[, knn_probability_columns])
summary(rowSums(knn_probabilities))
```

Frequently used annotation columns include:

```{r}
annotation_columns <- c(
  "cell_type_hard",
  "core_group",
  "knn_label",
  "knn_confidence"
)
annotation_columns[annotation_columns %in% names(cd)]
```

The next vignette demonstrates marker-aware tree gating and visualisation. The
third vignette uses high-confidence cells as a supervised-learning reference.

# Session information

```{r}
sessionInfo()
```
