---
title: "Machine-Learning Prediction"
author: "cytoGateR authors"
vignette: >
  %\VignetteIndexEntry{3. Machine-Learning Prediction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Overview

Tree gating can identify high-confidence reference cells while leaving
ambiguous cells unassigned. This vignette uses the stored `core_group` labels
to train a weighted k-nearest-neighbour classifier and predict the remaining
cells.

The example uses one complete image and reduced cross-validation settings to
remain suitable for package checks.

```{r}
library(cytoGateR)
data("cytoGateR_example", package = "cytoGateR")

image_ids <- SummarizedExperiment::colData(cytoGateR_example)$image_name
spe <- cytoGateR_example[, image_ids == image_ids[1L]]
table(SummarizedExperiment::colData(spe)$core_group)
```

# Train a weighted kNN reference

Only cells with a known `core_group` label are used during reference cleaning.
The `"Unassigned"` cells are retained for subsequent prediction.

```{r}
knn_reference <- train_custom_knn(
  spe,
  label_col = "core_group",
  assay_name = "exprs",
  unknown_label = "Unassigned",
  features = c("CD3e", "CD20", "CD4", "CD8a"),
  cv_folds = 2,
  repeats = 1,
  agreement_thresh = 0.5,
  k = 3,
  method = "pearson",
  seed = 1,
  BPPARAM = BiocParallel::SerialParam()
)
```

The returned object contains a cleaned reference and cross-validated
probabilities for retained reference cells:

```{r}
names(knn_reference)
dim(knn_reference$model$reference_data)
table(knn_reference$model$reference_labels)
```

# Predict unassigned cells

```{r}
knn_prediction <- predict_unknown_with_knn(
  knn_reference$spe,
  knn_reference,
  assay_name = "exprs",
  label_col = "cleaned_core_label",
  out_col = "knn_label_filled",
  pred_col = "knn_pred",
  unknown_label = "Unassigned",
  threshold = 0.5,
  k = 3,
  dist_method = "pearson"
)
```

```{r}
prediction_cd <- SummarizedExperiment::colData(knn_prediction$spe)
table(prediction_cd$knn_label_filled)
summary(prediction_cd$knn_confidence)
```

# Assemble the full probability matrix

`train_custom_knn()` returns probabilities for retained reference cells, while
`predict_unknown_with_knn()` returns probabilities for predicted cells. They
can be combined and aligned to the cells present in the updated object.

```{r}
full_prob_mat <- rbind(
  knn_reference$core_prob_mat,
  knn_prediction$prob_mat
)
full_prob_mat <- full_prob_mat[colnames(knn_prediction$spe), , drop = FALSE]

dim(full_prob_mat)
summary(rowSums(full_prob_mat))
```

# Quantify uncertainty

The uncertainty calculation combines probability-distribution measures with a
sample-aware spatial-discordance score.

```{r}
uncertainty <- calculate_uncertainty(
  full_prob_mat,
  knn_prediction$spe,
  sample_col = "sample_id",
  k_spatial = 3
)
head(uncertainty)
```

# Compare labels

```{r}
label_confusion_matrix(
  knn_prediction$spe,
  label_col1 = "cell_type_hard",
  label_col2 = "knn_label_filled"
)
```

```{r}
plot_labelled_cells(
  knn_prediction$spe,
  col_label = "knn_label_filled",
  image_col = "image_name"
)
```

# Other classifiers

`cytoGateR` also provides random-forest and optional Torch-based deep-learning
classifiers. They use the same core-label and unknown-cell workflow. The
following production alternatives are not evaluated during vignette building.

```{r}
#| eval: false
rf_reference <- train_custom_randomforest(
  spe,
  label_col = "core_group",
  unknown_label = "Unassigned",
  cv_folds = 5,
  repeats = 10,
  seed = 1
)

dl_reference <- train_custom_dl(
  spe,
  label_col = "core_group",
  unknown_label = "Unassigned",
  epochs = 50,
  cv_folds = 5,
  repeats = 3,
  seed = 1
)
```

# Session information

```{r}
sessionInfo()
```
