---
title: "Tree Gating and Visualisation"
author: "cytoGateR authors"
vignette: >
  %\VignetteIndexEntry{2. Tree Gating and Visualisation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Overview

This vignette demonstrates an interpretable tree-gating workflow on one
complete image from `cytoGateR_example`. A small two-lineage marker table keeps
the build fast while exercising the same functions used by larger analyses.

```{r}
library(cytoGateR)
data("cytoGateR_example", package = "cytoGateR")

image_ids <- SummarizedExperiment::colData(cytoGateR_example)$image_name
spe <- cytoGateR_example[, image_ids == image_ids[1L]]
spe
```

# Define a lineage table

Each row defines a cell type using positive and negative protein markers.
Marker columns are list-columns so that each cell type may use a different
number of markers.

```{r}
lineage_table <- data.frame(
  cell_type = c("Tcell", "Bcell"),
  pos_markers = I(list("CD3e", "CD20")),
  neg_markers = I(list("CD20", "CD3e"))
)
lineage_table
```

# Fit tree-gating models

For package-build speed, the example limits each tree to one split. Larger
analyses can increase `max_depth` and use more detailed marker definitions.

```{r}
tree_result <- run_tree_gating(
  spe,
  lineage_table,
  assay_name = "exprs",
  max_depth = 1,
  min_cells = 100,
  min_score = 0.1,
  uncert_thresh = 0.25,
  workers = 1
)
```

The result contains the updated object, fitted trees, probability matrix, hard
labels, and cleaned lineage table:

```{r}
names(tree_result)
head(tree_result$prob_mat)
table(tree_result$hard_label)
```

# Inspect the fitted trees

```{r}
print_celltype_tree(tree_result$trees$Tcell)
plot_celltype_tree(tree_result$trees$Tcell)
```

# Select high-confidence cells

`apply_cutoff_labels()` applies a cell-type-specific probability cutoff. A cell
receives a reference label only when exactly one cell type passes its cutoff.

```{r}
core_result <- apply_cutoff_labels(
  tree_result,
  cutoff_fn = function(x) stats::quantile(x, 0.95, na.rm = TRUE),
  label_col = "core_group",
  unknown_label = "Unknown"
)

table(SummarizedExperiment::colData(core_result$spe)$core_group)
```

# Visualise expression and annotations

Marker distributions help assess whether the selected markers separate the
expected populations.

```{r}
plot_marker_density(
  core_result$spe,
  marker = "CD3e",
  assay_name = "exprs",
  image_col = "image_name"
)
```

The gated labels can also be displayed in spatial coordinates:

```{r}
plot_labelled_cells(
  core_result$spe,
  col_label = "core_group",
  image_col = "image_name"
)
```

# Visualise stored full-workflow probabilities

The packaged object contains probabilities from the full ten-lineage demo
workflow. These provide a richer example for probability visualisation without
rerunning the complete model during vignette construction.

```{r}
plot_probability_map(
  spe,
  cell_type = "Bcell",
  image_col = "image_name"
)
```

```{r}
plot_label_counts(
  spe,
  label_col = "knn_label"
)
```

# Session information

```{r}
sessionInfo()
```
